### 2022-11-18 tex/semirings.tex
[ edit ] some corrections and notes after meeting

The methods described in \cref{sec:simple} for simply typed $\lambda$-calculus
make crucial use of \emph{weakening} --- the fact that if we have
$\Gamma \vdash A$, then we also have $\Gamma, \Delta \vdash A$.
We use this property to update environments as they go under binders.
However, as we saw in \cref{sec:linearity}, there are interesting calculi in
which general weakening does not hold.
As such, one of the aims of this chapter will be to find a form of weakening
applicable to variables of any type, while essentially retaining linearity
(as opposed to affinity).

As explained by \citet{McBride16}, the first insight is to, instead of
removing variables from the context of certain subterms, add an annotation to
free variables saying whether they are to be used or not.
We may use an annotation $\gr0$ on variables that are not to be used, and an
annotation $\gr1$ on variables that are to be used.
This convention lets us transcribe the usual $\otimes$-introduction rule
(left) as a rule with usage annotations (right).
I use lowercase $\gamma$ and $\delta$ to emphasise that the whole context now
looks like $\grP\gamma$, where $\grP$ is a list of usage annotations
$\gr{r_1}, \ldots, \gr{r_m}$ and $\gamma$ is a list of types $A_1, \ldots, A_m$
of the same length.
Explicit contexts will usually be written with usage annotations and types
interspersed, as $\gr{r_1}A_1, \ldots, \gr{r_m}A_m$.
I use $\gr r\gamma$ to abbreviate $\gr rx_1, \ldots, \gr rx_m$.

@@

What the existence of $\gr0$ has bought us is that variables never go out of
scope in subterms; rather, we lose the ability to use certain variables that
remain in scope.
Additionally, we recover a form of weakening: if $\Gamma \vdash A$, then also
$\Gamma, \gr0\delta \vdash A$, because the resulting term indeed uses no
variables from $\delta$.
A rigorous proof of weakening for terms will come in \cref{sec:lrsub}.

@@

To conceptualise the criteria on the usage annotations involved in this rule,
I introduce a formal $+$ operation on usage annotations.
The rule stated above relies on the facts that $\gr1 + \gr0 = \gr1$,
$\gr0 + \gr1 = \gr1$, and $\gr\omega + \gr\omega = \gr\omega$.
Addition lifts pointwise to vectors of usage annotations (the green capital
calligraphic $\grP$, $\grQ$, and $\grR$).
It is worth noting now that, thanks to the fact that $\gr0 + \gr0 = \gr0$, the
rule on the right below now accepts $\gr0$-annotated variables in its
conclusion, which is essential for weakening to be admissible.

@@

Some other transcriptions are as follows.
I unify the variable rules (the one for linear variables and the one for
intuitionistic variables) by introducing a coercibility ordering $\leq$ on usage
annotations.
We have $\gr\omega \leq \gr1$ because an intuitionistic variable can fill the
demand of a linear variable by dereliction.
We also have $\gr\omega \leq \gr0$, because intuitionistic variables can be
weakened away like $\gr0$-annotated variables.
All together, this means that at the unified variable rule, the variable being
used must have annotation less than or equal to $\gr1$, and every other variable
must have annotation less than or equal to $\gr0$.
Thus the basis vector $\langle x \rvert$ describes the upper bound on the
current usage context $\grR$.

@@

The final interesting rule form to cover is that found in DILL's
$\oc$-introduction rule.
I choose to see DILL's $\oc$-introduction as an $\gr\omega$-ary counterpart to
$\otimes$-introduction, though with the same premise each time rather than
$\gr\omega$-many premises.
This explanation recovers the behaviour that only $\gr\omega$- and
$\gr0$-annotated variables can appear in the conclusion of $\oc$-introduction,
and also justifies the designation of multiplication (vector scaling) as the
algebraic operation controlling the $\oc$-modality.

@@

In summary, the structure we have required of the set of usage annotations is
that they have addition (for $\otimes$-introduction and similar rules),
multiplication (for $\oc$-introduction), a $1$ (for a variable being used), a
$0$ (for a variable being discarded), and an ordering (allowing for subsumption
of usage restrictions).
Together, these form a \emph{partially ordered semiring} (posemiring), the laws
of which are both supported by examples and necessary for the syntax to be well
behaved.
\todo{Refer back to preliminaries for definition of posemiring}
For concreteness, I collect together the definition of the
$\{\gr0, \gr1, \gr\omega\}$ posemiring I have been using so far.

@@

\section{A usage-annotated calculus}\label{sec:lr}
\input{lr.tex}
\section{Bunched connectives}\label{sec:lnd}
\input{lnd.tex}
\section{What are linear renaming and substitution?}\label{sec:lrkits}
\input{lrkits.tex}
\section{Properties of linear environments}\label{sec:lenv}
\input{lenv.tex}
\section{Substitution is admissible in \name{}}\label{sec:lrsub}
\input{lrsub.tex}
\section{Adding recursion to \name{}}\label{sec:rec}
\input{rec.tex}
\section{Addendum: (lack of) partiality}\label{sec:part}
\input{part.tex}

>>>>

The methods described in \cref{sec:simple} for simply typed $\lambda$-calculus
make crucial use of \emph{weakening} --- the fact that if we have
$\Gamma \vdash A$, then we also have $\Gamma, \Delta \vdash A$.
We use this property to update environments as they go under binders.
However, as we saw in \cref{sec:linearity}, there are interesting calculi in
which general weakening does not hold.
As such, one of the aims of this chapter will be to find a form of weakening
applicable to variables of any type, while essentially retaining linearity
(as opposed to affineness).

As explained by \citet{McBride16}, the first insight is to, instead of
removing variables from the context of certain subterms, add an annotation to
free variables saying whether or not they are to be used.
I use an annotation $\gr0$ on variables that are not to be used, and an
annotation $\gr1$ on variables that are to be used.
This convention lets us transcribe the usual $\otimes$-introduction rule
(below left) as a rule with usage annotations (below right).
In the notation on the right, I let $\Gamma = \grP\gamma$ and
$\Delta = \grQ\delta$, where $\Gamma$ is a whole context comprising a
\emph{usage context} $\grP$ and a \emph{typing context} $\gamma$.
A usage context is a list of usage annotations, so
$\grP = \gr{r_1}, \ldots, \gr{r_m}$ and a typing context is a list of types, so
$\gamma = A_1, \ldots, A_m$.
When combined, the usage context and the typing context will be of the same
length.
Explicit contexts will usually be written with usage annotations and types
interspersed, as $\gr{r_1}A_1, \ldots, \gr{r_m}A_m$.
I use $\gr r\gamma$ to abbreviate $\gr rx_1, \ldots, \gr rx_m$.

@@

The use of $\gr0$ gives us the property that variables never go out of
scope in subterms; rather, we lose the ability to use certain variables, but
retain the ability to refer to them metatheoretically.
Additionally, we recover a form of weakening: if $\Gamma \vdash A$, then also
$\Gamma, \gr0\delta \vdash A$, because the resulting term indeed uses no
variables from $\delta$.
I prove the admissibility of weakening for terms will come in \cref{sec:lrsub}.

@@

To conceptualise the criteria on the usage annotations involved in this rule,
I introduce an additive structure over usage annotations.
The rule stated above relies on the facts that $\gr1 + \gr0 = \gr1$,
$\gr0 + \gr1 = \gr1$, and $\gr\omega + \gr\omega = \gr\omega$.
Addition lifts pointwise to vectors of usage annotations (the green capital
calligraphic $\grP$, $\grQ$, and $\grR$).
It is worth noting now that, thanks to the fact that $\gr0 + \gr0 = \gr0$, the
rule on the right below now accepts $\gr0$-annotated variables in its
conclusion, which is essential for weakening to be admissible.

@@

Some other transcriptions are as follows.
I unify the variable rules (the one for linear variables and the one for
intuitionistic variables) by introducing a coercibility ordering $\leq$ on usage
annotations.
We have $\gr\omega \leq \gr1$ because an intuitionistic variable can fill the
demand of a linear variable by dereliction.
We also have $\gr\omega \leq \gr0$, because intuitionistic variables can be
weakened away like $\gr0$-annotated variables.
All together, this means that at the unified variable rule, the variable being
used must have annotation less than or equal to $\gr1$, and every other variable
must have annotation less than or equal to $\gr0$.
I write this requirement as $\grR \leq \langle x \rvert$, where
$\langle x \rvert$ is the \emph{basis vector} at position $x$.

@@

The final interesting rule form to cover is found in DILL's
$\oc$-introduction rule.
DILL's $\oc$-introduction can be though of as an $\gr\omega$-ary counterpart to
$\otimes$-introduction, though with the same premise each time rather than
$\gr\omega$-many premises.
This explanation recovers the behaviour that only $\gr\omega$- and
$\gr0$-annotated variables can appear in the conclusion of $\oc$-introduction,
and also justifies the designation of multiplication (vector scaling) as the
algebraic operation controlling the $\oc$-modality.

@@

In summary, the structure we have required of the set of usage annotations is
that they have addition (for $\otimes$-introduction and similar rules),
multiplication (for $\oc$-introduction), a $1$ (for a variable being used), a
$0$ (for a variable not being used), and an ordering (allowing for subsumption
of usage restrictions).
Together, these form a \emph{partially ordered semiring} (posemiring), the laws
of which are both supported by examples and necessary for the syntax to be well
behaved.
\todo{Refer back to preliminaries for definition of posemiring}
For concreteness, I collect together the definition of the
$\{\gr0, \gr1, \gr\omega\}$ posemiring I have been using so far.

@@

\section{A usage-annotated calculus $\name$}\label{sec:lr}
\input{lr.tex}
\section{Bunched connectives}\label{sec:lnd}
\input{lnd.tex}
\section{What are linear renaming and substitution?}\label{sec:lrkits}
\input{lrkits.tex}
\section{Properties of linear environments}\label{sec:lenv}
\input{lenv.tex}
\section{Substitution is admissible in \name{}}\label{sec:lrsub}
\input{lrsub.tex}
\section{Adding recursion to \name{}}\label{sec:rec}
\input{rec.tex}
\section{Addendum: (lack of) partiality}\label{sec:part}
\input{part.tex}
